# Program & Institutional Context {#context}

## The University of Memphis

The University of Memphis was founded under the auspices of the General Education Bill, enacted by the Tennessee Legislature in 1909. Known originally as West Tennessee Normal School, the institution opened its doors September 10, 1912. In 1925 the name of the college changed to West Tennessee State Teachers College. The college changed names again in 1941, becoming Memphis State College, and in 1950 graduate studies were initiated. In 1957 the state legislature designated Memphis State full university status. The first doctoral programs began in 1966. In 1983 MSU became the first public university in Tennessee to gain accreditation of its entire curriculum. In 1994 MSU became The University of Memphis, and the Ned R. McWherter Library was completed.\
\
Moving into the 21st century, the Kemmons Wilson School of Hospitality and Resort Management and the FedEx Institute of Technology have made their mark on the University landscape. Today, the University of Memphis is one of Tennessee's three comprehensive doctoral-extensive institutions of higher learning. Situated in a beautiful park-like setting in the state's largest city, it was the flagship of the Tennessee Board of Regents system; in 2016-2017 the UM transitioned to its own governing board. The university awards more than 3,000 degrees annually. With an enrollment of approximately 22,000 students, The University of Memphis has 24 Chairs of Excellence, more than any other Tennessee university, and five state-approved Centers of Excellence.\
\
**Mission:**\
The University of Memphis is a learner-centered metropolitan research university providing high quality educational experiences while pursuing new knowledge through research, artistic expression, and interdisciplinary and engaged scholarship.\
\
**Accreditation:**\
The University of Memphis is accredited by the Commission on Colleges of the Southern Association of Colleges and Schools to award bachelor's, first professional, master's, educational specialist, and doctoral degrees.\
\
**Major Divisions of the University:**\
College of Arts and Sciences, Fogelman College of Business and Economics, College of Communication and Fine Arts, College of Education, Herff College of Engineering, Kemmons Wilson School of Hospitality and Resort Management, University College, Loewenberg College of Nursing, School of Communication Sciences and Disorders, School of Public Health, School of Health Sciences, Cecil C. Humphreys School of Law, and the Graduate School.\
\
**Degrees:**\
The University of Memphis offers 17 bachelor's degrees in more than 250 areas of study, master's degrees in over 55 subjects, and doctoral degrees in 23 disciplines in addition to the Juris Doctor (law) and a specialist degree in education.\
\
**Library:**\
The University's Ned R. McWherter Library provides one of the most electronically up-to-date information repositories within hundreds of miles. Library collections contain more than 13 million items, which include monographs, periodical volumes, federal and state documents, maps, and manuscripts. Holdings include nearly 3.1 million microformat materials and more than a million bound volumes.\
\
**Information Technology.**\
The U of M provides state-of-the-art computing facilities for student and faculty use, including a 24-hour TigerLAN lab. Thirty-five additional labs with more than 700 PC and Macintosh workstations and nearly "smart" classrooms complement teaching and research.\
\
**Center and Chairs of Excellence:**\
The University has five state-approved Centers of Excellence: Center for Applied Psychological Research, Center for Earthquake Research and Information, Center for Research Initiatives and Strategies for the Communicatively Impaired, Center for Research in Educational Policy, and the Institute of Egyptian Art and Archaeology. There are 25 endowed Chairs of Excellence.

## University Data {#unidata}

The University of Memphis operates a large, urban campus with a strong faculty and administrative base, supported by substantial annual resources.

-   **Operating Budget:** \$478 million annually
-   **Full-Time Faculty:** Over 1,000
-   **Total Employees:** Approximately 2,500 (faculty and staff combined)
-   **Campus Size:** 1,160 acres

### Faculty Demographic Data

Table \@ref(tab:facdemo) summarizes the demographic composition of faculty by race, gender, and appointment type for the 2024 academic year. Data are organized separately for female and male faculty, with categories for non-tenure track, tenure track, and tenured appointments. A small number of visiting faculty (n=2) are excluded from the table.

```{r facdemo, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(scales)

# Original faculty data by Race and Gender
faculty_data <- data.frame(
  Race = rep(c("Asian", "Black", "Hispanic", "Multi Race", "Other", "White"), times = 2),
  Gender = rep(c("Female", "Male"), each = 6),
  NonTenure = c(11, 63, 6, 6, 2, 201,
                 22, 21, 5, 2, 2, 108),
  TenureTrack = c(19, 18, 4, 2, 0, 56,
                   24, 10, 5, 1, 1, 53),
  Tenured = c(21, 14, 10, 1, 1, 111,
              46, 16, 13, 4, 1, 175)
)

# Calculate totals
faculty_data <- faculty_data %>%
  mutate(Total = NonTenure + TenureTrack + Tenured)

# Add 'Not Reported' row 
faculty_not_reported <- data.frame(
  Race = "Asian",
  Gender = "Not Reported",
  NonTenure = 2,
  TenureTrack = 0,
  Tenured = 0,
  Total = 2
)

faculty_data <- bind_rows(faculty_data, faculty_not_reported)

# Group and summarize female, male, and not reported
faculty_female <- faculty_data %>%
  filter(Gender == "Female") %>%
  bind_rows(
    summarise(., Race = "Total", Gender = "Female",
              NonTenure = sum(NonTenure), TenureTrack = sum(TenureTrack),
              Tenured = sum(Tenured), Total = sum(Total))
  )

faculty_male <- faculty_data %>%
  filter(Gender == "Male") %>%
  bind_rows(
    summarise(., Race = "Total", Gender = "Male",
              NonTenure = sum(NonTenure), TenureTrack = sum(TenureTrack),
              Tenured = sum(Tenured), Total = sum(Total))
  )

faculty_unknown <- faculty_data %>%
  filter(Gender == "Not Reported") %>%
  bind_rows(
    summarise(., Race = "Total", Gender = "Not Reported",
              NonTenure = sum(NonTenure), TenureTrack = sum(TenureTrack),
              Tenured = sum(Tenured), Total = sum(Total))
  )

# Combine for final display
faculty_split <- bind_rows(faculty_female, faculty_male, faculty_unknown)

# Format and display table
faculty_split %>%
  mutate(across(c(NonTenure, TenureTrack, Tenured, Total), ~ comma(.))) %>%
  select(-Gender) %>%
  kbl(
     caption = "Faculty Demographics by Appointment Type",
     col.names = c("Race", "Non-Tenure Track", "Tenure Track", "Tenured", "Total"),
    align = "lrrrr",
    format = "html"
  ) %>%
  add_header_above(c(" " = 1, "Appointment Type" = 3, " " = 1)) %>%
  group_rows("Female", 1, nrow(faculty_female)) %>%
  group_rows("Male", nrow(faculty_female) + 1, nrow(faculty_female) + nrow(faculty_male)) %>%
  group_rows("Not Reported", nrow(faculty_female) + nrow(faculty_male) + 1, nrow(faculty_split)) %>%
  kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    font_size = 13
  )
```

**Note.** This table presents full-time faculty by gender, race, and appointment type. Visiting and Ineligible faculty (n = 5) were excluded from this summary.\
\
\
**Tuition & Fees for Graduate Students (2025-2025 data; Full-Time, 10+ Credit Hours):** 
- In-State: \$6,780/year\
- Out-of-State: \$9,140/year\
- International: \$11,500/year

**Enrollment & Student Demographics**

As of Fall 2024, the University of Memphis serves over 20,000 students across undergraduate and graduate programs.

-   **Total Enrollment:** \~20,272
    -   Undergraduate: \~15,766
    -   Graduate: \~3,093

**Average ACT Score (Entering Freshmen):** 
- University of Memphis: **23**\
- Tennessee Average: **19.1**\
- U.S. Average: **20.3**

### Undergraduate and Graduate Student Demographic Data

The tables below summarize the race and gender distribution of undergraduate and graduate students at the university for the 2024 academic year. Table \@ref(tab:ugraddemo) presents demographic data for undergraduate students, while Table \@ref(tab:graddemo) provides the same breakdown for graduate students. Each table includes headcounts and percentages by gender and race, offering a clear snapshot of the student population’s composition. These data help contextualize the diversity represented within the university community.\
\

```{r ugraddemo, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(dplyr)
library(scales)

# Undergraduate Data
ug_data <- data.frame(
  Race = c("Black", "White", "Hispanic", "Asian", "Multi Race", "Non Resident Alien",
           "Not Specified", "Am Indian/Alaskan", "Native Hawaiian/Pacific Isl."),
  Female_Headcount = c(3490, 2421, 909, 298, 337, 126, 32, 10, 8),
  Female_Percent = c(45.73, 31.73, 11.91, 3.91, 4.42, 1.65, 0.42, 0.13, 0.10),
  Male_Headcount = c(1695, 2061, 637, 306, 213, 139, 44, 5, 5),
  Male_Percent = c(33.20, 40.37, 12.48, 5.99, 4.17, 2.72, 0.86, 0.10, 0.10),
  Total_Headcount = c(5185, 4482, 1546, 604, 551, 265, 76, 15, 13),
  Total_Percent = c(40.71, 35.19, 12.14, 4.74, 4.33, 2.08, 0.60, 0.12, 0.10)
)

ug_totals <- ug_data %>%
  summarise(
    Race = "Total",
    Female_Headcount = sum(Female_Headcount),
    Female_Percent = 100.00,
    Male_Headcount = sum(Male_Headcount),
    Male_Percent = 100.00,
    Total_Headcount = sum(Total_Headcount),
    Total_Percent = 100.00
  )

ug_data_full <- bind_rows(ug_data, ug_totals) %>%
  mutate(across(ends_with("Headcount"), comma),
         across(ends_with("Percent"), ~ sprintf("%.2f", .)))

ug_data_full %>%
  kbl(
  caption = "Undergraduate Student Race and Gender",
  col.names = c("Race", "Female", "%", "Male", "%", "Total", "%"),
  align = "lrrrrrr",
  format = "html"
) %>%
add_header_above(c(" " = 1, "Headcount" = 6)) %>%
kable_styling(
  full_width = TRUE,
  bootstrap_options = c("striped", "hover", "condensed", "responsive"),
  font_size = 13
) %>%
row_spec(nrow(ug_data_full), bold = TRUE)

```

\
\

```{r graddemo, echo=FALSE, message=FALSE, warning=FALSE}

# Graduate Data
grad_data <- data.frame(
  Race = c("White", "Non Resident Alien", "Black", "Hispanic", "Asian",
           "Multi Race", "Am Indian/Alaskan", "Not Specified"),
  Female_Headcount = c(1013, 515, 867, 150, 99, 79, 7, 6),
  Female_Percent = c(37.01, 18.82, 31.68, 5.48, 3.62, 2.89, 0.26, 0.22),
  Male_Headcount = c(621, 718, 267, 59, 58, 31, 5, 6),
  Male_Percent = c(35.14, 40.63, 15.11, 3.34, 3.28, 1.75, 0.40, 0.34),
  Total_Headcount = c(1636, 1233, 1134, 209, 157, 110, 12, 12),
  Total_Percent = c(36.31, 27.36, 25.17, 4.64, 3.48, 2.44, 0.27, 0.27)
)

grad_totals <- grad_data %>%
  summarise(
    Race = "Total",
    Female_Headcount = sum(Female_Headcount),
    Female_Percent = 100.00,
    Male_Headcount = sum(Male_Headcount),
    Male_Percent = 100.00,
    Total_Headcount = sum(Total_Headcount),
    Total_Percent = 100.00
  )

grad_data_full <- bind_rows(grad_data, grad_totals) %>%
  mutate(across(ends_with("Headcount"), comma),
         across(ends_with("Percent"), ~ sprintf("%.2f", .)))

grad_data_full %>%
  kbl(
    caption = "Graduate Student Race and Gender",
    col.names = c("Race", "Female", "%", "Male", "%", "Total", "%"),
    align = "lrrrrrr",
  format = "html") %>%
  add_header_above(c(" " = 1, "Headcount" = 6)) %>%
  kable_styling(
  full_width = TRUE,
  bootstrap_options = c("striped", "hover", "condensed", "responsive"),
  font_size = 13) %>%
  row_spec(nrow(grad_data_full), bold = TRUE)
```

\
\
**University Maintained Data & Dashboards:**\
This section provides links to University of Memphis maintained fact sheets and dashboards that may provide more up-to-date information about the University of Memphis as well as University of Memphis faculty, employee, and student data. 
- [UofM Facts at a Glance](https://www.memphis.edu/about/facts.php)\
- [Faculty Statistics](https://www.memphis.edu/oir/data/public_faculty.php)\
- [Employee Headcount](https://www.memphis.edu/oir/data/public_employee_headcount.php)\
- [Enrollment Trends](https://www.memphis.edu/oir/data/public_enrollment_trends.php)\
- [Student Demographics](https://www.memphis.edu/oir/data/public_student_demographics.php)

------------------------------------------------------------------------

## The Department of Psychology

**Mission:** Psychology is a major discipline which contributes to the social and natural sciences. The primary mission of the Department of Psychology is to advance the science and profession of psychology through the production and dissemination of knowledge related to the discipline, and the preparation of academic and professional psychologists. Consistent with the mission of the Center for Applied Psychological Research (CAPR), departmental emphasis is placed upon scholarly activities which maintain and enhance the department’s national research reputation. The department strives to provide high quality liberal and general education for undergraduate students, a coherent, high-quality program for its majors and graduate students, service courses to students in other majors, and services to the general public. The breadth of the faculty’s commitment is reflected in graduate programs in both basic and applied science and in diverse research and scholarly pursuits. The department’s mission is consistent with the university’s mission to advance learning through excellence in teaching, research, and service. In 2020-2021 the department employed 22 full-time and 10 Non-Tenure-Track faculty (including adjuncts), and 24 full-time office staff including the Institute for Intelligent Systems (IIS) and the Psychological Services Center (PSC).

**History and Degrees:** Psychology was established as an independent department in 1947. Authorization to award the master's degree was granted in 1960 and the first such degree was granted in 1962. The Ph.D. program in psychology was begun in 1966. For fall semester 2021 there were 112 students enrolled: There were 40 doctoral students in clinical, 12 in experimental, and 15 in school psychology. There were 29 students enrolled in the Master of Science in General Psychology (MSGP) and 16 in the Master of Arts/Educational Specialist in School Psychology (MA/EdS) Program. Graduate students have a voice in policy and decision-making through elected representatives to the department, and students are full members of several committees in the department. The training program in clinical psychology has been continuously accredited by the American Psychological Association since 1972. The MA/EdS program in school psychology has been continuously NASP approved since 1988. The PhD school psychology program was accredited by the APA in fall, 2014. More detailed history about the program is available in Chapter \@ref(history).

**Facilities:** The Department of Psychology occupies a building of 88,000 square feet on the main University campus. This four-story structure contains two complete floors of animal and human research laboratories allowing all types of psychological investigation, a clinic (Psychological Services Center), undergraduate laboratories, a data analysis center, classrooms, and offices for all Psychology Department faculty and full-time graduate students. Classrooms are modern and equipped with computer, projection equipment, and internet access. The department's laboratories and clinic are outfitted completely with modern equipment sufficient for faculty or student research or training activities. Additional space exists in the nearby FedEx Institute of Technology Bldg.

The University maintains a modern computer center with numerous networked connections in the Department of Psychology. Student access to this system is readily available. The department owns many network-connected computers. Wireless internet access is available to faculty and students in all campus instructional buildings. The department also houses two computer laboratories that are used for both research and teaching. Online and distance learning facilities have expanded to meet the needs of continued instruction and business operations during the pandemic.

**The Psychological Services Center (PSC).** The PSC at the University of Memphis offers a wide range of services to the university and the public. The Center is the training and service delivery unit of the Psychology Department’s graduate programs in clinical and school psychology. Psychological and psychoeducational assessment, consultation, and intervention services are available by graduate students and advanced Ph.D. candidates who are directly supervised by clinical and school psychologists from the Memphis community and the University faculty. Center of Excellence. Founded in 1984, the Center for Applied Psychological Research (CAPR) is one of the original Tennessee Centers of Excellence, now designated an Accomplished Center of Excellence. The CAPR faculty have attracted millions of dollars in extramural funding to the State of Tennessee, published hundreds of scholarly works, graduated hundreds of masters and doctoral students and many times that number of undergraduates, been elected or appointed to many national professional offices, and won numerous local and national honors and awards. The primary CAPR goals are (a) to achieve national prominence for its accomplishments, (b) to increase the number of trained psychologists able to conduct, disseminate, and apply such research, and (c) to secure extramural funding to facilitate research and training. In addition, as part of a Psychology Department in a major urban research university, CAPR supports the general teaching, research, and service missions of the department and university. Center funds are also used to sponsor major conferences that allow top psychologists from this country and throughout the world to visit our department. The primary mission of the Center for Applied Psychological Research (CAPR) at The University of Memphis is to support psychological research. As part of a major urban research university, a primary focus of this mission is the application of psychological knowledge to finding solutions to individual and social problems -- local, regional, and national. Therefore, CAPR research focuses extensively on problems concerning health, mental health, education, schools, crime, the environment, and children and their families. In addition, to increasing our capacity for generating new approaches to applied problems, CAPR also supports basic scientific research in cognitive psychology, biopsychology, social psychology, developmental psychology, and research design and statistics.

**Collaboration:** Faculty and graduate students conduct research and work on collaborative service projects in numerous community agencies, including the Desoto County Schools, Shelby County Schools, the municipal school districts of Arlington, Bartlett, Collierville, Germantown, Lakeland, and Millington, area charter schools and the Achievement School Districts including the ASPIRE Schools; Shelby County Head Start, the Boling Developmental Disabilities Center, the Veterans Administration Medical Center, the Departments of Psychiatry, Community Medicine, Pediatrics and Preventive Medicine at the University of Tennessee-Center for the Health Sciences-Memphis, St. Jude Children’s Research Hospital, the Memphis Mental Health Institute, agencies and departments of the City of Memphis, the University's Center for Student Development and Disability Resources for Students, and several community agencies. Administrators of these agencies are receptive to making their facilities and subject populations available for diverse psychological research projects and clinical services. Several distinguished scholars in psychology and related areas visit the department each year to present colloquia and workshops to faculty and students. The speakers represent a variety of current subject matter areas within psychology and have contributed to the Department’s Diversity Series held across the academic year. Additionally, the local chapters of Sigma Xi and other campus and community groups often invite scholars in psychology and related disciplines to Memphis.

**Research Areas:** The faculty, students, and programs are spread over 5 research areas: Behavioral Neuroscience; Child and Family Studies; Clinical Health Psychology; Cognitive Psychology; and Psychotherapy Research. All faculty and students in the school psychology programs are affiliated with the Child and Family Studies area.

*The Child, Adolescent, and Family Studies (CAFS) Research Area.* CAFS includes faculty and students from several department programs including Child Clinical Psychology, Developmental Psychology, and School Psychology and represents all three of the department's doctoral degree concentrations (Clinical, Experimental, and School). The common core of faculty research interests is a commitment to examining age-related cognitive, cultural, emotional, health, and social factors across infancy, childhood, and adolescence. Basic and applied research projects are conducted in a variety of settings, including community agencies, day care and preschool programs, homes, hospitals, and schools. From this diversity across faculty interests, graduate students can construct individualized programs of study and be involved with a wide variety of research topics and populations. CFS faculty have teaching assignments to courses related to the research area, and many provide supervised practicum experiences through the department's Psychological Services Center. Colloquium meetings are scheduled as needed on Friday afternoons during which faculty, students, and community partners present research findings. [Sources: Psychology Website, Undergraduate and Graduate Bulletins, Admissions Secretary, Governance Policy Manual.] The College of Education Teacher education has been a strong mission since the University was founded as a normal school in 1912.

## The College of Education

**The College of Education (COE)** is dedicated to preparing students for careers in a variety of professional and related occupational areas. Three academic departments make up the College of Education: Counseling, Educational Psychology and Research; Instruction and Curriculum Leadership; and Leadership. One department offers undergraduate degrees, and all departments offer a variety of graduate degree programs. School psychology students take courses in several of these departments including the areas of counseling, educational psychology and research, educational leadership, reading, and special education.

The COE also has several research and service units: Advising and Student Services, The Center for the Study of Higher Education, The Reading Center, The Center for Rehabilitation and Employment Research, The Center for Research in Educational Policy, The Institute on Disability, and The Barbara K. Lipman Early Childhood School and Research Institute.

## Conceptual Framework: Preparing Educational Leaders

The conceptual framework of the University of Memphis’ College of Education consists of its vision, mission, and core values. We have high expectations for candidate performance. It ensures that we are engaged with our community, relevant to our stakeholders and disciplines, and leading our profession.

The vision of the College of Education is driven by a commitment to diversity, social justice and equity. The College of Education is a premier educational organization that engages in innovative and impactful research, teaching, and service.

The mission of the College of Education is to build on a foundation of success in research, teaching, and service by:

-   fostering depth and breadth of knowledge conducting relevant and innovative research
-   developing culturally competent leaders and practitioners
-   collaborating with and providing services to diverse and local, national, and international partners to address real world problems of practice
-   advancing an environment where diversity, social justice, and equity are paramount.

The core values of the College of Education include:

-   *Diversity* – We recognize, include and embrace all individuals ensuring intercultural relationships.
-   *Inclusion* – We actively and intentionally commit to promoting equity and social justice in every endeavor.
-   *Respect* – We value all people, recognizing that good ideas can come from anyone, anywhere, at any time.
-   *Innovation* – We stay on the cutting edge by identifying new ways to embrace technology and solve problems.
-   *Service* – We utilize our skills and attributes to enhance our practices and expand our support to others.
-   *Build* on the foundation of success in research, teaching and service by fostering depth and breadth of knowledge, conducting relevant and innovative research, developing culturally competent leaders and practitioners.

[Back to top](#program-handbook)

<small>*Last updated: `r format(file.info("sppc.Rmd")$mtime, "%B %d, %Y")`*</small>
